The Rockerverse: Packages and Applications for Containerisation with R

### 2020-03-04 rockerverse.tex
fix some widows and orphans
https://en.wikipedia.org/wiki/Widows_and_orphans

Docker is not the only containerisation software. An alternative
stemming from the domain of high-perfomance computing is
\textbf{Singularity} \citep{kurtzer_singularity_2017}. Singularity can
run Docker images. Rocker images work out of the box if the main process
is R, e.g., in \texttt{rocker/r-base}, but Singularity does not succeed
in running images where there is an init script, e.g., in containers
that by default run RStudio~Server. In the latter case, a
\texttt{Singularity} file, a recipe akin to a \texttt{Dockerfile}, needs
to be used to make necessary adjustments. To date, no comparable image
stack to the Rocker Project's images exists on
\href{https://singularity-hub.org/}{Singularity Hub}. A further tool for
running containers is
\href{https://github.com/containers/libpod}{\textbf{podman}}, which can
also build \texttt{Dockerfile}s and run Docker images. Proof of concepts
exists for using podman to build and run Rocker
containers\footnote{See \href{https://github.com/nuest/rodman}{https://github.com/nuest/rodman} and \href{https://github.com/rocker-org/rocker-versioned/issues/187}{https://github.com/rocker-org/rocker-versioned/issues/187}},
but the prevelance of Docker, especially in the broader user community
beyond experts or niche systems and the vast amount of blog posts and
courses for Docker currently cap specific development efforts for both
Singularity and podman in the R community. This might quickly change if
the usability and spread of Singularity or podman increase, or if
security features such as rootless/unprivileged containers, which both
these tools support out of the box, become more sought after.

@@

The \emph{Bioconductor} \texttt{Dockerfile} selects the desired version
of R from Rocker, adds required system dependencies, and uses the
\CRANpkg{BiocManager} package for installing appropriate versions of
\emph{Bioconductor} packages \citep{cran_biocmanager}. A strength of
this approach is that the responsibility for complex software
configuration (including customized development) is shifted from the
user to the experienced \emph{Bioconductor} core team. However, a recent
audit of the \emph{Bioconductor} image stack \texttt{Dockerfile} led to
the deprecation of several community-maintained images, because the
numerous specific images became too hard to understand, complex to
maintain, and cumbersome to customise. As part of the simplification, a
recent innovation is to produce a \texttt{bioconductor\_docker:devel}
image to emulate the \emph{Bioconductor} nightly Linux build machine as
closely as possible. This image contains the build system environment
variables and the \emph{system dependencies} needed to install and check
almost all (1813 out of 1823) \emph{Bioconductor} software packages, and
it saves users and package developers from managing these themselves.
Furthermore, the image is configured so that \code{.libPaths()} has
\samp{/usr/local/lib/R/host-site-library} as the first location. Users
mounting a location on the host file system to this location can
persistently manage installed packages across Docker containers or image
updates. Many R users pursue flexible work flows tailored to particular
analysis needs rather than standardized work flows. The new
\texttt{bioconductor\_docker} image is well suited for this preference,
while \texttt{bioconductor\_docker:devel} provides developers with a
test environment close to \emph{Bioconductor}'s build system.

@@

\textbf{\CRANpkg{liftr}} aims to solve the problem of persistent
reproducible reporting in statistical computing based on the R Markdown
format \citep{xie2018} for dynamic documents
\citep[\href{https://nanx.me/liftr/}{https://nanx.me/liftr/}, ][]{liftr2019}.
The irreproducibility of authoring environments can become an issue for
collaborative documents and large-scale platforms for processing
documents. \pkg{liftr} makes the document the main and sole workflow
control file and the only file that needs to be shared between
collaborators for consistent environments, e.g.~demonstrated in the
DockFlow project (\url{https://dockflow.org}). It introduces new fields
to the R Markdown document header, allowing users to manually declare
the dependencies, including versions, for rendering the document. The
package then generates a \texttt{Dockerfile} from this metadata and
provides a utitility function to render the document inside a Docker
container, i.e., \code{render\_docker("foo.Rmd")}. An RStudio addin even
allows compilation of documents with the single push of a button.

@@

Containers can also serve as useful playgrounds to create environments
ad hoc or to provide very specific environments that are not needed or
not easily available in day-to-day development for the purposes of
developing R packages. These environments may have specific versions of
R, of R extension packages, and of system libraries used by R extension
packages, and all of the above in a specific combination.

First, such containers can greatly facilitate \textbf{fixing bugs and
quick evaluation}, because developers and users can readily start and
later discard a container to investigate a bug report or try out a piece
of software \citep[cf.][]{ooms_opencpu_2017}. The container does not
affect their regular system. Using the Rocker images with RStudio, these
disposable environments lack no development comfort
(cf.~Section~\nameref{compendia}). \citet{ooms_opencpu_2017} describes
how \texttt{docker\ exec} can be used to get a root shell in a container
for customisation during software evaluation.
\citet{eddelbuettel_debugging_2019} describes an example of how a Docker
container was used to debug an issue with a package only occurring with
a particular version of Fortran, and using tools which are not readily
available on all platforms (e.g., not on macOS).

@@

The prevalence of Docker in industry naturally leads to the use of
containers with R in such settings as well, as companies already manage
platforms in Docker containers. These products often entail a large
amount of open source software in combination with proprietary layers
adding the relevant commercialisation features. One such example is
RStudio's data science platform
\href{https://rstudio.com/products/team/}{RStudio Team}. It allows teams
of data scientists and their respective IT/DevOps groups to develop and
deploy code in R and Python around the RStudio Open Source Server inside
of Docker images, without requiring users to learn new tools or directly
interact with containers. The best practices for
\href{https://support.rstudio.com/hc/en-us/articles/360021594513-Running-RStudio-with-Docker-containers}{running
RStudio with Docker containers} as well as
\href{https://github.com/rstudio/rstudio-docker-products}{Docker images}
for RStudio's commercial products are publicly available.

@@

The AI @ T-Mobile team created a set of neural network machine learning
natural language processing models to help customer care agents manage
text-based messages for customers \citep{t-mobile_enterprise_2018}. For
example, one model quickly identifies whether a message is from a
customer
\citep[see \pkg{Shiny}-based \href{https://secure.message.t-mobile.com/v1/shiny/is-customer/app/}{demo} further described by ][]{nolis_small_2019},
and others tell which customers are likely to make a repeat purchase. If
a data scientist creates a machine learning model and exposes it through
a \pkg{plumber} API, then someone else on the marketing team could write
software that sends different emails depending on that real-time
prediction. The models are convolutional neural networks that use the
\CRANpkg{keras} package \citep{cran_keras} and run in a Rocker
container. The corresponding \texttt{Dockerfile}s are published
\href{https://github.com/tmobile/r-tensorflow-api}{on GitHub}. Since the
models power tools for agents and customers, they need to have extremely
high uptime and reliability. The AI @ T-Mobile team found that the
models performed well, and today these models power real-time services
that are called over a million times a day.

@@

This article is a snapshot of the R corner in a universe of applications
built with a many-faced piece of software, Docker. \texttt{Dockerfile}s
and Docker images are the go-to methods for collaboration between roles
in an organisation, such as development and IT operations teams, and
between parties in the communication of knowledge, such as research
workflows or education. Docker has become synonymous with applying the
concept of containerisation to solve challenges of reproducible
environments, e.g., in research and in development \& production, and of
scalable deployments as it can easily move processing between machines
(e.g., locally, a cloud provider's VM, another cloud provider's
Container-as-a-Service). Reproducible environments, scalability and
efficiency, and portability across platforms/infrastructures are the
common themes behind R packages, use cases, and applications in this
work.

The R packages and use cases presented above show the growing number of
users, developers, and real-world applications in the community and the
resulting innovations. But the applications also point to the challenges
of keeping up with a continuously evolving landscape. The use cases
contributed by co-authors also have a degree of overlap, which can be
expected as a common language and understanding of good practices is
still taking shape. Also, the ease with which one can create a complex
software systems with Docker, such as an independent Docker image stack,
to serve one's specific needs leads to parallel developments. This
ease-of-DIY in combination with the difficulty of reusing parts from or
composing multiple \texttt{Dockerfile}s is a further reason for
\textbf{fragmentation}. Instructions can be outsourced into
distributable scripts and then copied into the image during build, but
that makes \texttt{Dockerfile}s harder to read. Scripts added to a
\texttt{Dockerfile} also add a layer of complexity and increase the risk
of incomplete recipes. Despite the different image stacks presented
here, the pervasiveness of Rocker images can be traced back to its
maintainers and the user community valuing collaboration and shared
starting points over impulses to create individual solutions. Aside from
that, fragmentation may not be a bad sign but may instead be a
reflection of a growing market that is able to sustain multiple related
efforts. With the maturing of core building blocks, such as the Rocker
Project's suite of images, more systems will be built successfully, but
they may also be built behind the curtains. Docker alone, as a flexible
core technology, is not a feasible level of collaboration and
abstraction. Instead, the use cases and applications observed in this
work provide a more useful division.

@@

The ordering of authors following DN and DE is alphabetical. DN
conceived the article idea,
\href{https://github.com/nuest/rockerverse-paper/issues/3}{initialised the formation of the writing team},
wrote sections not mentioned below, and revised all sections. DE wrote
the introduction and the section about containerisation and the Rocker
Project, and reviewed all sections. DB wrote the section on
\pkg{outsider}. GD contributed the CARD.com use case. RC \& EH
contributed to the section on interfaces for Docker in R
(\emph{dynverse} and \texttt{dynwrap}). DC contributed content on
Gigantum. ME contributed to the section on processing and deployment to
cloud services. CF wrote paragraphs about \pkg{r-online},
\pkg{dockerfiler}, \pkg{r-ci} and \pkg{r-db}. LK contributed content on
\pkg{dockr}. SL contributed content on RStudio's usage of Docker. BM
wrote the section on research compendia and made the project Binder
ready. HN \& JN co-wrote the section on the T-Mobile use case. KR wrote
the section about \pkg{holepunch}. NR wrote paragraphs about shared work
environments and GPUs. LS \& NT wrote the section on Bioconductor. PS
wrote the paragraphs about GitLab CI/CD pipelines with Shinyproxy and
OpenFaaS templates. TS \& JW wrote the section on CyVerse. CvP wrote the
section on the usage of Docker containers in Dodona. CW wrote the
sections on Whole Tale and contributed content about publication
reproducibility audits. NX contributed content on \pkg{liftr}. All
authors approved the final version. This articles was collaboratively
written at
\href{https://github.com/nuest/rockerverse-paper/}{https://github.com/nuest/rockerverse-paper/}.
The
\href{https://github.com/nuest/rockerverse-paper/graphs/contributors}{contributors page},
\href{https://github.com/nuest/rockerverse-paper/commits/master}{commit history},
and
\href{https://github.com/nuest/rockerverse-paper/issues/}{discussion issues}
provide a detailed view on the respective contributions.

>>>>

Docker is not the only containerisation software. \textbf{Singularity}
stems from the domain of high-perfomance computing
\citep{kurtzer_singularity_2017} and can also run Docker images. Rocker
images work out of the box if the main process is R, e.g., in
\texttt{rocker/r-base}, but Singularity does not succeed in running
images where there is an init script, e.g., in containers that by
default run RStudio~Server. In the latter case, a \texttt{Singularity}
file, a recipe akin to a \texttt{Dockerfile}, needs to be used to make
necessary adjustments. To date, no comparable image stack to the Rocker
Project's images exists on
\href{https://singularity-hub.org/}{Singularity Hub}. A further tool for
running containers is
\href{https://github.com/containers/libpod}{\textbf{podman}}, which can
also build \texttt{Dockerfile}s and run Docker images. Proof of concepts
exists for using podman to build and run Rocker
containers\footnote{See \href{https://github.com/nuest/rodman}{https://github.com/nuest/rodman} and \href{https://github.com/rocker-org/rocker-versioned/issues/187}{https://github.com/rocker-org/rocker-versioned/issues/187}},
but the prevelance of Docker, especially in the broader user community
beyond experts or niche systems and the vast amount of blog posts and
courses for Docker currently cap specific development efforts for both
Singularity and podman in the R community. This might quickly change if
the usability and spread of Singularity or podman increase, or if
security features such as rootless/unprivileged containers, which both
these tools support out of the box, become more sought after.

@@

The \emph{Bioconductor} \texttt{Dockerfile} selects the desired R
version from Rocker images, adds required system dependencies, and uses
the \CRANpkg{BiocManager} package for installing appropriate versions of
\emph{Bioconductor} packages \citep{cran_biocmanager}. A strength of
this approach is that the responsibility for complex software
configuration and customization is shifted from the user to the
experienced \emph{Bioconductor} core team. However, a recent audit of
the \emph{Bioconductor} image stack \texttt{Dockerfile} led to the
deprecation of several community-maintained images, because the numerous
specific images became too hard to understand, complex to maintain, and
cumbersome to customise. As part of the simplification, a recent
innovation is the \texttt{bioconductor\_docker:devel} image, which
emulates the \emph{Bioconductor} environment for nightly builds as
closely as possible. This image contains the environment variables and
the system dependencies needed to install and check almost all
\emph{Bioconductor} software packages (1813 out of 1823). It saves users
and package developers from creating this environment themselves.
Furthermore, the image is configured so that \code{.libPaths()} has
\samp{/usr/local/lib/R/host-site-library} as the first location. Users
mounting a location on the host file system to this location can
persistently manage installed packages across Docker containers or image
updates. Many R users pursue flexible work flows tailored to particular
analysis needs rather than standardized work flows. The new
\texttt{bioconductor\_docker} image is well suited for this preference,
while \texttt{bioconductor\_docker:devel} provides developers with a
test environment close to \emph{Bioconductor}'s build system.

@@

\textbf{\CRANpkg{liftr}} \citep{liftr2019} aims to solve the problem of
persistent reproducible reporting in statistical computing based on the
R Markdown format \citep{xie2018}. The irreproducibility of authoring
environments can become an issue for collaborative documents and
large-scale platforms for processing documents. \pkg{liftr} makes the
dynamic R Markdown document the main and sole workflow control file and
the only file that needs to be shared between collaborators for
consistent environments, e.g.~demonstrated in the DockFlow project
(\url{https://dockflow.org}). It introduces new fields to the document
header, allowing users to manually declare the versioned dependencies
required for rendering the document. The package then generates a
\texttt{Dockerfile} from this metadata and provides a utitility function
to render the document inside a Docker container, i.e.,
\code{render\_docker("foo.Rmd")}. An RStudio addin even allows
compilation of documents with the single push of a button.

@@

Containers can also serve as playgrounds and provide specific or ad hoc
environments for the purposes of developing R packages. These
environments may have specific versions of R, of R extension packages,
and of system libraries used by R extension packages, and all of the
above in a specific combination.

First, such containers can greatly facilitate \textbf{fixing bugs and
code evaluation}, because developers and users can readily start a
container to investigate a bug report or try out a piece of software
\citep[cf.][]{ooms_opencpu_2017}. The container can later be discarded
and does not affect their regular system. Using the Rocker images with
RStudio, these disposable environments lack no development comfort
(cf.~Section~\nameref{compendia}). \citet{ooms_opencpu_2017} describes
how \texttt{docker\ exec} can be used to get a root shell in a container
for customisation during software evaluation without writing a
\texttt{Dockerfile}. \citet{eddelbuettel_debugging_2019} describes an
example of how a Docker container was used to debug an issue with a
package only occurring with a particular version of Fortran, and using
tools which are not readily available on all platforms (e.g., not on
macOS).

@@

The prevalence of Docker in industry naturally leads to the use of R in
containers, as companies already manage platforms in Docker containers.
These products often entail a large amount of open source software in
combination with proprietary layers adding the relevant
commercialisation features. One such example is RStudio's data science
platform \href{https://rstudio.com/products/team/}{RStudio Team}. It
allows teams of data scientists and their respective IT/DevOps groups to
develop and deploy code in R and Python around the RStudio Open Source
Server inside of Docker images, without requiring users to learn new
tools or directly interact with containers. The best practices for
\href{https://support.rstudio.com/hc/en-us/articles/360021594513-Running-RStudio-with-Docker-containers}{running
RStudio with Docker containers} as well as
\href{https://github.com/rstudio/rstudio-docker-products}{Docker images}
for RStudio's commercial products are publicly available.

@@

The AI @ T-Mobile team created a set of machine learning models for
natural language processing to help customer care agents manage
text-based messages from customers \citep{t-mobile_enterprise_2018}. For
example, one model identifies whether a message is from a customer
\citep[see \pkg{Shiny}-based \href{https://secure.message.t-mobile.com/v1/shiny/is-customer/app/}{demo} further described by ][]{nolis_small_2019},
and others tell which customers are likely to make a repeat purchase. If
a data scientist creates a such a model and exposes it through a
\pkg{plumber} API, then someone else on the marketing team can write
software that sends different emails depending on that real-time
prediction. The models are convolutional neural networks that use the
\CRANpkg{keras} package \citep{cran_keras} and run in a Rocker
container. The corresponding \texttt{Dockerfile}s are published
\href{https://github.com/tmobile/r-tensorflow-api}{on GitHub}. Since the
models power tools for agents and customers, they need to have extremely
high uptime and reliability. The AI @ T-Mobile team found that the
models performed well, and today these models power real-time services
that are called over a million times a day.

@@

This article is a snapshot of the R corner in a universe of applications
built with a many-faced piece of software, Docker. \texttt{Dockerfile}s
and Docker images are the go-to methods for collaboration between roles
in an organisation, such as developers and IT operators, and between
participants in the communication of knowledge, such as researchers or
students. Docker has become synonymous with applying the concept of
containerisation to solve challenges of reproducible environments, e.g.,
in research and in development \& production, and of scalable
deployments because it can easily move processing between machines,
e.g., locally, a cloud provider's VM, another cloud provider's
Container-as-a-Service. Reproducible environments, scalability \&
efficiency, and portability across infrastructures are the common themes
behind R packages, use cases, and applications in this work.

The projects presented above show the growing number of users,
developers, and real-world applications in the community and the
resulting innovations. But the applications also point to the challenges
of keeping up with a continuously evolving landscape. Some use cases
have considerable overlap, which can be expected as a common language
and understanding of good practices is still taking shape. Also, the
ease with which one can create a complex software systems with Docker to
serve one's specific needs, such as an independent Docker image stack,
leads to parallel developments. This ease-of-DIY in combination with the
difficulty of reusing parts from or composing multiple
\texttt{Dockerfile}s is a further reason for \textbf{fragmentation}.
Instructions can be outsourced into distributable scripts and then
copied into the image during build, but that makes \texttt{Dockerfile}s
harder to read. Scripts added to a \texttt{Dockerfile} also add a layer
of complexity and increase the risk of incomplete recipes. Despite the
different image stacks presented here, the pervasiveness of Rocker
images can be traced back to its maintainers and the user community
valuing collaboration and shared starting points over impulses to create
individual solutions. Aside from that, fragmentation may not be a bad
sign but may instead be a reflection of a growing market that is able to
sustain multiple related efforts. With the maturing of core building
blocks, such as the Rocker suite of images, more working systems will be
built, but they may simply work behind the curtains. Docker alone, as a
flexible core technology, is not a feasible level of collaboration and
abstraction. Instead, the use cases and applications observed in this
work provide a more useful division.

@@

The ordering of authors following DN and DE is alphabetical. DN
conceived the article idea,
\href{https://github.com/nuest/rockerverse-paper/issues/3}{initialised the formation of the writing team},
wrote sections not mentioned below, and revised all sections. DE wrote
the introduction and the section about containerisation and the Rocker
Project, and reviewed all sections. DB wrote the section on
\pkg{outsider}. GD contributed the CARD.com use case. RC \& EH
contributed to the section on interfaces for Docker in R
(\emph{dynverse} and \texttt{dynwrap}). DC contributed content on
Gigantum. ME contributed to the section on processing and deployment to
cloud services. CF wrote paragraphs about \pkg{r-online},
\pkg{dockerfiler}, \pkg{r-ci} and \pkg{r-db}. LK contributed content on
\pkg{dockr}. SL contributed content on RStudio's usage of Docker. BM
wrote the section on research compendia and made the project Binder
ready. HN \& JN co-wrote the section on the T-Mobile use case. KR wrote
the section about \pkg{holepunch}. NR wrote paragraphs about shared work
environments and GPUs. LS \& NT wrote the section on Bioconductor. PS
wrote the paragraphs about GitLab CI/CD pipelines with Shinyproxy and
OpenFaaS templates. TS \& JW wrote the section on CyVerse. CvP wrote the
section on the usage of Docker containers in Dodona. CW wrote the
sections on Whole Tale and contributed content about publication
reproducibility audits. NX contributed content on \pkg{liftr}. All
authors approved the final version. This articles was collaboratively
written at
\href{https://github.com/nuest/rockerverse-paper/}{https://github.com/nuest/rockerverse-paper/}.
The
\href{https://github.com/nuest/rockerverse-paper/graphs/contributors}{contributors page}
and
\href{https://github.com/nuest/rockerverse-paper/issues/}{discussion issues}
provide details on the respective contributions.
